%
%  revised  front.tex  2011-09-02  Mark Senn  http://engineering.purdue.edu/~mark
%  created  front.tex  2003-06-02  Mark Senn  http://engineering.purdue.edu/~mark
%
%  This is ``front matter'' for the thesis.
%
%  Regarding ``References'' below:
%      KEY    MEANING
%      PU     ``A Manual for the Preparation of Graduate Theses'',
%             The Graduate School, Purdue University, 1996.
%      TCMOS  The Chicago Manual of Style, Edition 14.
%      WNNCD  Webster's Ninth New Collegiate Dictionary.
%
%  Lines marked with "%%" may need to be changed.
%

  % Dedication page is optional.
  % A name and often a message in tribute to a person or cause.
  % References: PU 15, WNNCD 332.
\begin{dedication}
To Siva, who always believed it is better to go down trying.
\end{dedication}

  % Acknowledgements page is optional but most theses include
  % a brief statement of apreciation or recognition of special
  % assistance.
  % Reference: PU 16.
\begin{acknowledgments}

First and foremost, I would like to thank my advisor Suresh Jagannathan, who
graciously took a naive yet enthusiastic graduate student into his research
group, and taught him how to do research. I am immensely thankful for his
always open office door, uncanny ability to foresee potential issues, high
standards and patience. Without his guidance and support, this dissertation
would not have been possible. I have been fortunate to have Patrick Eugster,
Jan Vitek, and Dongyan Xu on my PhD committee. Their critiques and insightful
comments helped focus and refine the dissertation.

I would like to thank my colleague, mentor and good friend Lukasz Ziarek, with
whom I worked closely in my early years. He showed by example the mechanics of
going from an idea to a finished paper. The early collaborations with him set
the tone for the rest of the research in this dissertation. I had the good
fortune to study distributed systems over two graduate courses under Patrick
Eugster. His enthusiasm and industry inspired me to push for publishing the
final class project. I learnt from him the tact of fleshing out abstract ideas
and producing a polished product.

I have had the privilege of working with immensely talented people over my
internships. The summer spent at Samsung Research was a turning point for me in
no small part due to the encouragement and enthusiasm of Daniel Waddington.
Samsung not only funded the internship, but also supported some of my time at
graduate school through a generous gift, for which I am very grateful. The
internship at Microsoft Research, Cambridge broadened my research perspectives.
Tim Harris, Simon Marlow and Simon Peyton Jones taught me how to remain focused
and grounded.

I am indebted to the friendship and insight of my numerous friends at Purdue
University including Siddharth Narayanaswamy, Karthik Swaminathan Nagaraj,
Gowtham Kaki, K.R. Jayaram, Naresh Rapolu, Pawan Prakash, Gustavo Petri, He
Zhu, Siddharth Tiwary, Karthik Kambatla, Raghavendra Prasad, Armand Navabi, and
many others. I will miss you all.

I would like to thank my parents for their constant source of encouragement and
offering me complete freedom to pursue my interests. Never did they doubt my
completing this journey even when I doubted myself. Last but not least, my
greatest debt is to my wife, Siva, whose undying support during difficult times
gave me the strength to keep on fighting. No amount of gratitude can repay my
debt to you.
\end{acknowledgments}

  % The Table of Contents is required.
  % The Table of Contents will be automatically created for you
  % using information you supply in
  %     \chapter
  %     \section
  %     \subsection
  %     \subsubsection
  % commands.
  % Reference: PU 16.
\tableofcontents

  % If your thesis has tables, a list of tables is required.
  % The List of Tables will be automatically created for you using
  % information you supply in
  %     \begin{table} ... \end{table}
  % environments.
  % Reference: PU 16.
\listoftables

  % If your thesis has figures, a list of figures is required.
  % The List of Figures will be automatically created for you using
  % information you supply in
  %     \begin{figure} ... \end{figure}
  % environments.
  % Reference: PU 16.
\listoffigures

\begin{abbreviations}
	ACML& Asynchronous Concurrent ML\cr
	CC& Causally Consistent\cr
	CML& Concurrent ML\cr
	CSH& Cached Shared Heap\cr
	DMA& Direct Memory Access\cr
	EC& Eventually Consistent\cr
	ECDS& Eventually Consistent Data Store\cr
	GC& Garbage-Collector\cr
	MAV& Monotonic Atomic View\cr
	MPB& Message-Passing Buffers\cr
	RC& Read Committed\cr
	RDT& Replicated Data Type\cr
	RR& Repeatable Read\cr
	SC& Strongly Consistent\cr
  SCC& Single-chip Cloud Computer\cr
	SMC& Software-Managed Cache-coherence\cr
	USH& Uncached Shared Heap\cr
\end{abbreviations}

  % Abstract is required.
  % Note that the information for the first paragraph of the output
  % doesn't need to be input here...it is put in automatically from
  % information you supplied earlier using \title, \author, \degree,
  % and \majorprof.
  % Reference: PU 17.
\begin{abstract}

In recent years, there has been a wide-spread adoption of both multicore and
cloud computing. Traditionally, concurrent programmers have relied on the
underlying system providing \emph{strong memory consistency}, where there is a
semblance of concurrent tasks operating over a shared global address space,
with reads to a memory location witnessing the latest write to the same
location. However, providing scalable strong consistency guarantees as the
scale of the system grows is an increasingly difficult endeavor. In a
multicore setting, the increasing complexity and the lack of scalability of
hardware abstractions such as on-chip cache coherence deters scalable strong
consistency. In geo-distributed compute clouds, the latency and availability
concerns in the presence of partial failures prohibit strong consistency.
Hence, modern multicore and cloud computing platforms eschew strong consistency
in favor of weakly consistent memory, meaning that each task has a view of
memory that is incomparable with the other tasks. As a result, programmers on
these platforms must tackle the full complexity of concurrent programming for
an asynchronous distributed system.

This dissertation argues that functional programming language abstractions can
simplify scalable concurrent programming for weakly consistent systems.
Functional programming espouses mutation-free programming, and rare mutations
when present are explicit in their types. By controlling and explicitly
reasoning about shared state mutations, functional programming mitigates the
complexity of programming weakly consistent systems. Building upon this
intuition, this dissertation presents three major contributions, each focused
on addressing a particular challenge associated with weakly consistent loosely
coupled systems: the absence of cache coherence, asynchrony and eventual
consistency. First, it describes \MMSCC, a parallel extension of MLton Standard
ML compiler and runtime for the Intel Single-chip cloud computer (SCC), and
shows how to provide an efficient cache coherent virtual address space on top
of a \emph{non cache coherent} multicore architecture. Next, it describes
\rxcml, a distributed extension of \MM and shows that, with the help of
speculative execution, synchronous communication can be utilized as an
efficient abstraction for programming asynchronous distributed systems.
Finally, it presents \quelea, a programming system for \emph{eventually
consistent} distributed stores, and shows that the choice of correct
consistency level for replicated data type operations and transactions can be
automated with the help of high-level declarative \emph{contracts}.

\end{abstract}
